---
title: "Hair shedding joining and import"
author: "Harly Durbin"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: paged
    code_folding: show
header-includes:
  - \setlength{\parindent}{4em}
  - \setlength{\parskip}{0em}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(rlang)
library(lubridate)
library(ggplot2)
library(glue)
library(magrittr)
library(purrr)
library(readxl)
library(tidylog)

fescue_belt <- 
  read_excel(here::here("data/raw_data/fescue_coordinates.xlsx"))

is.wholenumber <- function(x, tol = .Machine$double.eps^0.5)  abs(x - round(x)) < tol

```

```{r, warning=FALSE, message=FALSE}
source(here::here("source_functions/hair_ped.R"))

```

# Notes & questions

* ~~Batch effects from non-integer data~~?
    + ~~Could try building in an effect for study/source~~?
    + Or does contemporary group take care of that?
* How is foot score treated? Docility? As continuous? 
* Changed `toxic_fescue` to `NO` for Texas scores
* Figure colors:
    + `#538797`: darker blue
    + `#C2D9CD`: lighter blue
    + `#FCAF1`: off-white
    + `#FF842A`: orange
    + `#FED55D`: yellow
---

* Age key based [here](http://www.angus.org/Nce/Documents/ByTheNumbers0108.pdf) and [BIF age of dam cutoffs](https://beefimprovement.org/wp-content/uploads/2013/07/BIFGuidelinesFinal_updated0916.pdf#23)

```{r}
# Had to do this programatically my brain was melting
age_key <- 
  tibble(class = 1:25) %>% 
  mutate(
    # The actual day you start being an age
    actual_start = class*365,
    # Mininum is higher for yearling
    min =
      case_when(
        class == 1 ~ 274,
        TRUE ~ actual_start - 90
      ),
    max = actual_start + 274,
    total_days = max - min
    ) %>% 
  select(age_class = class, min, actual_start, max, total_days)

```

```{r}
age_key
```


# Mizzou data

```{r}
animal_table <- read_rds(here::here("data/raw_data/animal_table.rds"))
```


* Pull records from Mizzou data where breed code is "AN" and/or listed breed association is "American Angus Association"
* Correct some registration numbers after searching in Pro Rodeo & find AAA registration for animals cross-registered in other breed associations
    + Drop hair scores from animals cancelled for parentage or not found

```{r}
mu_long <- 
  # Mizzou hair shedding project cleaned data
  read_rds(here::here("data/raw_data/cleaned.rds")) %>% 
    # janky registration corrections I did by searching Pro Rodeo
  left_join(animal_table %>%
              select(Lab_ID, Reg, breed_assoc, breed_assoc_reg)) %>%
  filter(breed_code == "AN" |
           breed_assoc == "American Angus Association") %>%
  filter(!farm_id %in% c("SAV", "BAT", "UMF")) %>%
  # janky registration corrections I did by searching Pro Rodeo
  left_join(
    read_excel(here::here("data/derived_data/reg_search.xlsx"), col_types = "text") %>%
      select(farm_id, animal_id, an_reg),
    by = c("farm_id", "animal_id"),
  ) %>%
  # Remove if cancelled for parentage or not found
  filter(an_reg != 0 | is.na(an_reg)) %>%
  mutate(
    # Pull from breed_assoc_reg (MU animal database)
    registration_number = case_when(
      # reg num found in Pro Rodeo
      # breed_assoc_reg from MU animal database
      # Diff breed reg, replace with Angus
      !is.na(an_reg) ~ an_reg,
      str_detect(breed_assoc_reg, "^1|^6") ~ breed_assoc_reg,
      # NA, replace with Angus
      is.na(registration_number) &
        str_detect(Reg, "^1|^6") ~ Reg,      
      TRUE ~ registration_number
    ),
    # Fix toxic_fescue, calving season, date score recorded for some farms
    toxic_fescue =
      case_when(
        is.na(toxic_fescue) & farm_id %in% c("OHF", "GOF", "PVF") ~ "YES",
        is.na(toxic_fescue) & farm_id %in% c("LAF") ~ "NO",
        TRUE ~ toxic_fescue
      ),
    calving_season = 
      case_when(
        farm_id %in% c("UMCT", "HAF", "GOF") ~ "SPRING", 
        TRUE ~ calving_season
    ), 
    date_score_recorded =
      case_when(
        is.na(date_score_recorded) & farm_id == "RCF" ~ ymd("2019-05-25"),
        TRUE ~ date_score_recorded
      ),
  ) %>% 
  # Remove if still no registration number
  filter(!is.na(registration_number)) %>% 
  # Remove if not an AAA or BIR reg
  filter(str_detect(registration_number, "^1|^6")) %>% 
  select(animal_id,
         farm_id,
         breed_code,
         registration_number,
         barcode,
         sex,
         date_score_recorded,
         hair_score,
         age,
         calving_season,
         toxic_fescue,
         year,
         Lab_ID) %>% 
  group_by(registration_number, date_score_recorded) %>%
  # Remove duplicates
  filter(n() == 1) %>% 
  ungroup()

```

# AGI Legacy data

```{r}
legacy_data <-
  # Raw Excel file provided by Mike MacNeil
  read_excel(here::here("data/raw_data/FinalFile(WW work)_MacNeil0619.xlsx"),
             na = "0") %>%
  janitor::clean_names() %>%
  # Remove empty rows and columns
  janitor::remove_empty(which = c("rows", "cols")) %>%
  select(
    farm_id = owner,
    location,
    registration_number = dam_reg,
    animal_id = cow_id,
    age = dam_bd,
    date_score_recorded = date,
    calf_dob = calf_bd,
    contains("hc"),
    region
  ) %>%
  # Convert from "wide" to "long" format:
  # One row per score-r per day
  reshape2::melt(
    id = c(
      "farm_id",
      "location",
      "registration_number",
      "animal_id",
      "age",
      "date_score_recorded",
      "calf_dob",
      "region"
    )
  ) %>%
  rename(scored_by = variable,
         hair_score = value) %>%
  mutate(
    scored_by = as.character(scored_by),
    registration_number = as.character(registration_number),
    scored_by = case_when(
      scored_by == "hcs" ~ "A",
      scored_by == "hc_sa" ~ "B",
      scored_by == "hc_sb" ~ "C"
    )
  ) %>%
  # Remove empty rows
  filter(!is.na(hair_score)) %>%
  # Impute toxic_fescue
  mutate(
    year = lubridate::year(date_score_recorded),
    date_score_recorded = ymd(date_score_recorded),
    toxic_fescue =
      case_when(
        # Presumably no hot fescue in TX
        region == "TX" ~ "NO",
        # In previous hair shedding paper, authors indicated that the entire NCSU Upper Piedmont research station herd grazes hot fescue
        farm_id %in% c("UPRS") ~ "YES",
        TRUE ~ NA_character_
      ),
    # Only cows scored in AGI data
    sex = "F",
    # Infer age from birth date and date scored
    age = as.integer(lubridate::time_length(
      ymd(date_score_recorded) - ymd(age), unit = "years"
    )),
    # Infer calving season from calf dob
    calving_season =
      case_when(
        between(
          lubridate::month(calf_dob),
          left = 1,
          right = 6
        ) ~ "SPRING",
        between(
          lubridate::month(calf_dob),
          left = 7,
          right = 12
        ) ~ "FALL"
      ),
    # collapse/rename some farm ids
    farm_id = 
      case_when(
        farm_id %in% c("QH") ~ "QuakerHill",
        farm_id %in% c("4") ~ "Creswell",
        farm_id %in% c("A") ~ "bradley3",
        farm_id %in% c("B") ~ "RobertElliott",
        farm_id %in% c("I") ~ "Woodall",
        farm_id %in% c("F") ~ "Fleming",
        TRUE ~ farm_id
      )
  ) %>% 
  group_by(registration_number, date_score_recorded) %>%
  # Take the mean of all scores for that day
  mutate(hair_score = mean(hair_score)) %>%
  ungroup() %>%
  select(-scored_by) %>%
  distinct()
```


# Bradley 3 data

```{r}

# Tidy Bradley 3 hair shedding data provided by James Henderson
bradley3_data <-
  readxl::read_excel(here::here("data/raw_data/henderson_hair_shed.xlsx"), 
           skip = 1) %>% 
  # Select all columns but name
  dplyr::select(
    animal_id = 1,
    registration_number = 3, 
    age = 4,
    hs19 = 5,
    hs18 = 6
    ) %>% 
  # Change from column for each year to row for each observation
  reshape2::melt(id = c("animal_id", "registration_number", "age")) %>% 
  dplyr::mutate(
    variable = case_when(
      variable == "hs19" ~ ymd("2019-05-01"),
      variable == "hs18" ~ ymd("2018-05-08")
      ),
    registration_number = as.character(registration_number)) %>% 
  dplyr::rename(
    hair_score = value,
    date_score_recorded = variable
    ) %>% 
  dplyr::mutate(
    farm_id = "bradley3",
    date_score_recorded = lubridate::ymd(date_score_recorded),
    year = lubridate::year(date_score_recorded),
    # Infer age from birth date and date scored
    age = as.integer(lubridate::time_length(
      date_score_recorded - lubridate::ymd(age), unit = "years"
      )),
    toxic_fescue = "NO",
    calving_season = "FALL"
    ) %>% 
  # Remove empty rows
  dplyr::filter(!is.na(hair_score)) 
```


# Join & tidy

* Remove animals where sex on file does not match sex in pedigree
* Remove scores from animals where age on file > 2 year different from age based on DOB in pedigree

```{r}
angus_join <-
bind_rows(
  mu_long %>%
    mutate(source = "mizzou"),
  legacy_data %>%
    mutate(source = "legacy"),
  bradley3_data %>%
    mutate(source = "bradley3")
) %>%
  # full_reg with BIR or AAA pre-pended
  mutate(full_reg =
           case_when(
             str_detect(registration_number, "^6") ~ as.character(glue("BIR{registration_number}")),
             str_detect(registration_number, "^1") ~ as.character(glue("AAA{registration_number}"))
           )) %>%
  # Use for finding likely errors
  left_join(hair_ped %>%
              rename(sex_ped = sex,
                     dob_ped = dob),
            by = c("full_reg")) %>%
  # Pull in missing calving seasons from by-hand ProRodeo searching
  left_join(read_excel(
    here::here("data/derived_data/season_search.xlsx"),
    trim_ws = TRUE,
    skip = 1,
    col_names = c("farm_id", "registration_number", "year", "cs2")
  )) %>%
  # Fix incorrect calving seasons
  # For missing calving season, impute using dob
  mutate(
    calving_season =
      case_when(
        !is.na(cs2) ~ cs2,
        sex == "F" &
          is.na(calving_season) &
          between(lubridate::month(dob_ped), left = 1, right = 6) ~ "SPRING",
        sex == "F" &
          is.na(calving_season) &
          between(lubridate::month(dob_ped), left = 7, right = 12) ~ "FALL",
        TRUE ~ calving_season
      ),
    # Remove spaces from farm ID
    farm_id = str_remove(farm_id, " "),
    # Infer age in days using date score recorded
    age_ped =
      case_when(
        !is.na(date_score_recorded) ~ as.integer(date_score_recorded - dob_ped),
        # If no date scored provided, use May 1 of that year
        is.na(date_score_recorded) ~ as.integer(ymd(glue::glue("{year}-05-01")) - dob_ped)
      )
  ) %>%
  filter(age_ped >= 274) %>%
  # Age classes based on BIF age of dam
  mutate(
    age_class =
      case_when(
        between(age_ped, 274, 639) ~ 1,
        between(age_ped, 640, 1004) ~ 2,
        between(age_ped, 1005, 1369) ~ 3,
        between(age_ped, 1370, 1734) ~ 4,
        between(age_ped, 1735, 2099) ~ 5,
        between(age_ped, 2100, 2464) ~ 6,
        between(age_ped, 2465, 2829) ~ 7,
        between(age_ped, 2830, 3194) ~ 8,
        between(age_ped, 3195, 3559) ~ 9,
        between(age_ped, 3560, 3924) ~ 10,
        between(age_ped, 3925, 4289) ~ 11,
        between(age_ped, 4290, 4654) ~ 12,
        between(age_ped, 4655, 5019) ~ 13,
        between(age_ped, 5020, 5384) ~ 14,
        between(age_ped, 5385, 5749) ~ 15,
        between(age_ped, 5750, 6114) ~ 16,
        between(age_ped, 6115, 6479) ~ 17,
        between(age_ped, 6480, 6844) ~ 18,
        between(age_ped, 6845, 7209) ~ 19,
        between(age_ped, 7210, 7574) ~ 20,
        between(age_ped, 7575, 7939) ~ 21
      ),
    keep =
      case_when(
        # Remove if sex on file differs from sex in pedigree
        sex != sex_ped ~ "drop sex",
        # Remove if age on file greater than 2 years different than age in pedigree
        abs(age - age_class) > 2 ~ "drop age",
        TRUE ~ "keep"
        ),
    age =
      case_when(
        is.na(age_class) ~ as.integer(age),
        TRUE ~ as.integer(age_class)
      ),
    sex =
      case_when(
        is.na(sex_ped) ~ sex,
        TRUE ~ sex_ped
      ),
  ) %>% 
  filter(!str_detect(keep, "drop")) %>% 
  select(
    farm_id,
    location,
    animal_id,
    full_reg, 
    registration_number,
    sex,
    date_score_recorded,
    year,
    hair_score,
    dob = dob_ped,
    age,
    calving_season,
    toxic_fescue,
    source, 
    Lab_ID,
    barcode
  ) %>% 
  # Add coordinates
  left_join(
    read_csv(here::here("data/raw_data/an_loc_key.csv"))
  ) %>% 
  distinct()
```
 
# Summary 

  
* **Result:** `r nrow(angus_join)` scores from `r n_distinct(angus_join$registration_number)` animals 

```{r, message=FALSE}
angus_join %>% 
    select(
    registration_number,
    sex,
    date_score_recorded,
    dob,
    age,
    calving_season,
    toxic_fescue
  ) %>% 
  visdat::vis_miss()
```

* Number of scores per farm

```{r, message=FALSE}
angus_join %>% 
  group_by(farm_id) %>% 
  tally(sort = TRUE)
```

* Summary by source

```{r, message=FALSE}
angus_join %>%
  mutate(source = if_else(source == "bradley3", "legacy", source)) %>%
  group_by(source) %>%
  summarise(n = n(),
            Mean = mean(hair_score),
            SD = sd(hair_score)) %>%
  bind_rows(
    angus_join %>%
      mutate(source = "All") %>%
      group_by(source) %>%
      summarise(
        n = n(),
        Mean = mean(hair_score),
        SD = sd(hair_score)
      )
  )
```

## Score distribution

* Distribution of scores (rounded to nearest whole number) in legacy & Bradley3 data

```{r, message= FALSE}
angus_join %>% 
  mutate(source = if_else(source == "bradley3", "legacy", source),
         hair_score2 = round(hair_score, 0)) %>%
  filter(source == "legacy") %>% 
  janitor::tabyl(hair_score2)
```

* Distribution of scores in Mizzou data

```{r, message=FALSE}
angus_join %>% 
  mutate(hair_score2 = round(hair_score, 0)) %>%
  filter(source == "mizzou") %>% 
  janitor::tabyl(hair_score2)
```

```{r, message=FALSE}
angus_join %>% 
  mutate(source = if_else(source == "bradley3", "legacy", source)) %>%
  ggplot(aes(x = hair_score, fill = source)) +
  geom_density(alpha = 0.3)
```

## Distribution of scoring timing across all years

```{r, message=FALSE}
angus_join %>% 
  mutate(date = ymd(format(angus_join$date_score_recorded, "2000-%m-%d"))) %>% 
  summarise(min(date, na.rm = TRUE),
            max(date, na.rm = TRUE))
```

```{r, message=FALSE}
angus_join %>% 
  mutate(date = ymd(format(angus_join$date_score_recorded, "2000-%m-%d"))) %>% 
  ggplot(aes(x = date)) +
  geom_density()
```

## Sex ratio

```{r, message=FALSE}
angus_join %>% 
  group_by(sex) %>% 
  tally()
```


## Location of scores

```{r, message=FALSE}
source(here::here("source_functions/hair_map.R"))

#cowplot::ggsave2(filename = here::here("figures/figure2.hair_map.pdf"), width = 85, height = 52.4, units = "mm", dpi = 300)

```

## Number of records per animal

```{r, message=FALSE}
angus_join %>% 
  group_by(full_reg) %>% 
  tally(sort = TRUE) %>%
  ungroup() %>% 
  group_by(n) %>% 
  summarise(n_animals = n()) %>% 
  ungroup() %>% 
  mutate(percent = scales::percent(n_animals/sum(n_animals)),
         n_animals = scales::comma(n_animals)) %>% 
  rename(n_records = n) 
```

# Export

```{r, eval=FALSE}
angus_join %>% 
  write_rds(here::here("data/derived_data/angus_join.rds")) 

```

# Notes and commentary

  Scores recorded on registered & BIR Angus cattle between 2016 and 2019 were extracted from the Mizzou Hair Shedding Project dataset. Records were dropped when the reported sex of an animal did not match the sex recorded in the AAA pedigree or and when the reported age differed from the age calculated using birth date recorded in the AAA pedigree for that registration by more than two years.  
  When no calving season was reported, it was imputed using the most recent natural birth calving date prior to the recorded score. When no natural birth calving dates were available, calving season was imputed using birth date.  
  In the AGI legacy project data, some animals were recorded by multiple scoring technicians on the same day. Where relevant, phenotypes recorded on the same animal on the same day were averaged. Fescue grazing status was not recorded in this data, but animals scored in Texas were assumed to not have grazed endophyte-infected fescue in spring of the scoring year.  
  This resulted in `r angus_join %>% filter(source == "mizzou") %>% n_distinct() %>% scales::comma()` scores recorded on `r angus_join %>% filter(source == "mizzou") %>% select(registration_number) %>% n_distinct(.)  %>% scales::comma()` animals in the Mizzou Hair Shedding Project and `r angus_join %>% filter(source %in% c("legacy", "bradley3")) %>% n_distinct() %>% scales::comma()` scores recorded on `r angus_join %>% filter(source %in% c("legacy", "bradley3")) %>% select(registration_number) %>% n_distinct(.)  %>% scales::comma()` in the AGI legacy project data.  
  