---
title: "F250 ID matching"
author: "Harly Durbin"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(rlang)
library(glue)
library(ggplot2)
library(lubridate)
library(forcats)
library(magrittr)
library(purrr)
library(tidylog)

source(here::here("source_functions/iterative_id_search.R"))
source(here::here("source_functions/coalesce_join.R"))

```


# Setup

> AGI combined hair shedding dataset

```{r}

angus_join <- 
  read_rds(here::here("data/derived_data/angus_join.rds")) %>% 
  mutate(
    Lab_ID =
      case_when(
        full_reg == "AAA16611219" ~ NA_integer_,
        TRUE ~ as.integer(Lab_ID)
      )
  )

```

> MU sample sheet

```{r, echo = TRUE, message=FALSE, warning=FALSE}

sample_sheet <- 
  read_delim(here::here("data/raw_data/190802_sample_sheet.csv"), delim = c(","), col_names = TRUE, guess_max = 10000) %>% 
  # This gets annoying so I'm renaming it
  rename(Lab_ID = lab_id,
         Reg = reg)

#write_rds(sample_sheet, here::here("data/raw/sample_sheet.rds"))
```

> MU animal table

```{r}
animal_table <- read_rds(here::here("data/raw_data/animal_table.rds"))
```

> Full MU hair shedding dataset 

```{r}
nested_join <- read_rds(here::here("data/raw_data/nested_join.rds"))

```

```{r}
geno_check <-
  angus_join %>%
  select(source, farm_id, Lab_ID, full_reg, registration_number, barcode) %>%
  distinct() %>%
  # Fill in barcodes by registration number
  group_by(full_reg) %>% 
  arrange(barcode) %>% 
  fill(barcode, .direction = "downup") %>% 
  ungroup() %>% 
  # Angus genotype check from Duc
  left_join(
    read_csv(
      here::here("data/raw_data/genotype_check.csv"),
      col_names = c("full_reg", "assay_agi"),
      na = "#N/A",
      col_types = cols(.default = "c")
    ) %>%
      mutate(
        full_reg =
          case_when(
            str_detect(full_reg, "^1") ~ glue("AAA{full_reg}"),
            str_detect(full_reg, "^6") ~ glue("BIR{full_reg}"),
            TRUE ~ full_reg
          )
      )
  ) %>%
  left_join(
    angus_join %>%
      filter(source == "mizzou") %>%
      select(source, farm_id, Lab_ID, full_reg, registration_number, barcode) %>%
      distinct() %>%
      # Add info from MU sample sheet
      left_join(sample_sheet %>%
                  rename(assay_mu = assay))
  ) %>%
  select(-contains("international_id")) %>%
  mutate(
    at_agi =
      case_when(!is.na(assay_agi) ~ "YES",
                TRUE ~ "NO"),
    at_mu =
      case_when(!is.na(assay_mu) ~ "YES",
                TRUE ~ "NO"),
    at_any =
      case_when(!is.na(assay_agi) | !is.na(assay_mu) ~ "YES",
                TRUE ~ "NO")
  ) %>%
  distinct()

```


# List of animals that should  have gentoypes for Duc


```{r}
f250_list <-
geno_check %>% 
  filter(assay_mu == "GGPF250") %>% 
  # Exclude samples not genotyped through the hair shedding project
  filter(date_added > lubridate::ymd("2016-08-01")) %>% 
  select(-do_not_analyze, -contains("assay"), -source, -farm_id, -registration_number, -Reg, -at_agi, -at_mu, -at_any) %>% 
  filter(!Lab_ID %in% c(142869)) 
```

> Export list of animals genotyped on F250 through MU 

```{r, eval = FALSE}

f250_list %>% 
  rename(registration_number = full_reg,
         lab_id = Lab_ID) %>%
  arrange(registration_number) %>% 
  write_csv(here::here("data/derived_data/20191114.F250.csv"))
```

> Export list of all animals in AGI combined dataset

```{r}
angus_join %>% 
  distinct(full_reg) %>% 
  write_csv(here::here("data/derived_data/20191114.all_individuals.csv"), col_names = FALSE)
```

# F250 sample IDs from Duc

* Duc: "Attached is a list 2698 unique sample IDs from my F250 genotypes. Some of your “lab_id” match my sample IDs. Some of your “sample_id” match my sample IDs too; and there are also many cases where both of your “lab_id” and “sample_id” match different samples, which create confusion. Therefore I think I better let you decide which of my sample IDs to be used."

```{r}

f250_id_search <-
read_table2(here::here("data/raw_data/191108_sample_ids.txt")) %>% 
  mutate(Lab_ID = NA_integer_) %>% 
  # Barcode in geno_check
  id_search(
    source_col = sampleID,
    search_col = barcode,
    search_df = geno_check,
    key_col = Lab_ID
    ) %>% 
  # Bob's sample_id in geno_check
  id_search(
    source_col = sampleID,
    search_col = sample_id,
    search_df = 
      geno_check %>% 
      mutate(sample_id = as.character(sample_id)),
    key_col = Lab_ID
    ) %>% 
  # Lab_ID in geno_check
  id_search(
    source_col = sampleID,
    search_col = labid2,
    search_df = 
      geno_check %>% 
      mutate(labid2 = as.character(Lab_ID)),
    key_col = Lab_ID
    ) %>%
  # AAA registration number from geno_check
  id_search(
    source_col = sampleID,
    search_col = full_reg,
    search_df = geno_check,
    key_col = Lab_ID
    ) %>%  
  # Lab_ID in animal_table
  id_search(
    source_col = sampleID,
    search_col = labid2,
    search_df = 
      animal_table %>% 
      mutate(labid2 = as.character(Lab_ID)),
    key_col = Lab_ID
    ) %>%
  # Ref_ID2 (usually barcode) in animal_table
  id_search(
    source_col = sampleID,
    search_col = Ref_ID2,
    search_df = animal_table,
    key_col = Lab_ID
    ) %>% 
  left_join(
    geno_check %>% 
      filter(!is.na(Lab_ID)) %>% 
      select(full_reg, Lab_ID)
    ) %>% 
  distinct() %>% 
  left_join(
    animal_table %>% 
      mutate(anreg = as.character(glue::glue("AAA{breed_assoc_reg}"))) %>% 
      select(Lab_ID, anreg)
  ) %>% 
  mutate(
    full_reg =
      case_when(
        is.na(full_reg) & !is.na(anreg) ~ anreg,
        TRUE ~ full_reg
      )
  ) %>% 
  select(-anreg)


```

> How many distinct registration numbers?

```{r}

f250_id_search %>% 
  filter(!is.na(Lab_ID)) %>% 
  distinct(full_reg)

```

> How many weren't in the list I sent Duc? What do they belong to?

```{r}
f250_id_search %>% 
  filter(!Lab_ID %in% f250_list$Lab_ID & !is.na(Lab_ID)) %>% 
  distinct(Lab_ID) %>% 
  left_join(animal_table) 
```

> How many have a duplicate?

```{r}
f250_id_search %>% 
  filter(!is.na(Lab_ID)) %>% 
  add_count(Lab_ID) %>% 
  filter(n > 1) %>% 
  arrange(Lab_ID) %>% 
  distinct(full_reg, n)
```

> How many not in MU data?

```{r}
f250_id_search %>% 
  filter(is.na(Lab_ID))
```


```{r, eval = FALSE}

f250_id_search %>% 
  arrange(order) %>% 
  rename(registration_number = full_reg,
         lab_id = Lab_ID) %>%
  write_csv(here::here("data/derived_data/20191120.F250_sample_ids.csv"), na = "")

```

```{r}
genotyped <-
  read_table2(here::here("data/derived_data/genotyped_id.txt"), col_names = FALSE) %>%
  pull(X1)
```

```{r}
f250_id_search %>% 
  filter(full_reg %in% genotyped) 
```

