---
title: "Breeding values & variance components"
author: "Harly Durbin"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(rlang)
library(glue)
library(ggplot2)
library(lubridate)
library(forcats)
library(magrittr)
library(readxl)
library(purrr)
library(tidylog)

source(here::here("source_functions/three_gen.R"))
source(here::here("source_functions/melt_aireml.R"))
source(here::here("source_functions/cg_tallies.R"))
source(here::here("source_functions/calculate_heritability.R"))
```

# Notes & questions

* ~~Write more about possibly why such low PE variance~~
* SNP file and map file formatting in `geno_format.Rmd`. 
* CG = farm, year, calving season (spring or fall), age group (yearling, 2-year-old, 3-year-old, or other), fescue, and score group
    + Remove CGs with fewer than 5 animals

## Threshold model?

* https://stats.stackexchange.com/questions/25894/changing-the-scale-of-a-variable-to-0-100

$$\frac{max_{new}-min_{new}}{max_{old}-min_{old}}*(v-max_{old})+max_{new} $$

```{r}

angus_join %>% 
  mutate(threshold = ((1-0)/(5-1))*(hair_score-5)+1) %>% 
  select(threshold, hair_score, everything())

```


# Setup

```{r, warning=FALSE, message=FALSE}
source(here::here("source_functions/hair_ped.R"))

```

```{r}
start <- 
  read_rds(here::here("data/derived_data/start.rds"))
```


```{r}
angus_join <-
  read_rds(here::here("data/derived_data/angus_join.rds"))
```


```{r}
ped <-
  start %>%
  left_join(hair_ped %>%
              select(full_reg, sire_reg, dam_reg))  %>%
  three_gen(full_ped = hair_ped) %>%
  mutate_all( ~ replace_na(., "0"))

```

# Threshold model?

https://stats.stackexchange.com/questions/25894/changing-the-scale-of-a-variable-to-0-100

$$\frac{max_{new}-min_{new}}{max_{old}-min_{old}}*(v-max_{old})+max_{new} $$

```{r}

angus_join %>% 
  mutate(threshold = ((1-0)/(5-1))*(hair_score-5)+1) %>% 
  select(threshold, hair_score, everything())

```



# Variance components, heritability, and repeatability

```{r}
var_cov <-
  melt_aireml(
    path = here::here("data/derived_data/general_varcomp/normal/airemlf90.general_varcomp.normal.log"),
    effect2 = c("hair_dir"),
    effect3 = c("hair_pe"),
    resids = c("hair_res")
  ) 
```

```{r}
var_cov %>% 
  filter(val1 == val2) %>% 
  separate(val1, into = c("key", "effect"), sep = "_") %>% 
  select(effect, var_cov) %>% 
  tidyr::pivot_wider(names_from = effect, values_from = var_cov) %>% 
  rename(`Genetic variance` = dir,
         `Permanent environment variance` = pe,
         `Residual variance` = res)
```



```{r}
herit <-
  var_cov %>%
  univ_heritability(
    abbrv = "hair",
    desc = "Hair shedding",
    pe = TRUE
    )

```

```{r}
herit
```


# Commentary


```{r, eval=FALSE}
read_table2(here::here("data/derived_data/general_varcomp/freqdata.count.after.clean"), col_names = FALSE) %>% 
  filter(X3 == "MAF") %>% 
  select(X1) %>% 
  n_distinct() %>% 
  scales::comma()


```

  Contemporary groups were defined as a combination of farm, year, calving season (spring or fall), fescue status, age group (yearling, 2-year-old, 3-year-old, or other), and score group. Score group was determined using a sliding window to maximize the number of animals per contemporary group in a seven-day window. In the future, it will be recommended that producers hair shedding score all cattle within a week of one another in order to enable accurate contemporary grouping. Contemporary groups with fewer than 5 animals were dropped. This resulted in `r start %>% n_distinct() %>% scales::comma()` total scores from `r start %>% select(registration_number) %>%  n_distinct(.) %>% scales::comma()` animals in `r start %>% select(cg_num) %>%  n_distinct(.)` contemporary groups.  

  The H-matrix was constructed using the 3-generation additive pedigree and imputed genotypes, resulting in `r ped %>% select(full_reg) %>% n_distinct(.) %>% scales::comma()` total animals (`r ped %>% select(sire_reg) %>% n_distinct(.) %>% scales::comma()` distinct sires, `r ped %>% select(dam_reg) %>% n_distinct(.) %>% scales::comma()` distinct dams).  

  Estimated narrow sense heritability ($\frac{\sigma^2_{a}}{\sigma^2_{a}+\sigma^2_{pe}+\sigma^2_{e}}$) was `r herit %>% pull(3)` and estimated repeatability ($\frac{\sigma^2_{a}+\sigma^2_{pe}}{\sigma^2_{a}+\sigma^2_{pe}+\sigma^2_{e}}$) was `r herit %>% pull(Repeatability)`.
