---
title: "Hair shedding score (single trait) SNP effects"
author: "Harly Durbin"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(rlang)
library(glue)
library(ggplot2)
library(lubridate)
library(forcats)
library(magrittr)
library(readxl)
library(purrr)
library(tidylog)

source(here::here("source_functions/hair_manhattan.R"))

```

# Notes & questions

* SNP file and map file formatting in `geno_format.Rmd`. 
* SNP effects calculated from ssBLUP using `blupf90` and `postGSf90`
* postGS output files explanation (p. 41)
    + (`windows_variance`)[http://nce.ads.uga.edu/wiki/doku.php?id=readme.pregsf90#misc_options]
        - 1: trait
        - 2: effect
        - 3: Start SNP number or SNP name for the window
        - 4: End SNP number or SNP name for the window
        - 5: window size (number of SNP)
        - 6: Start (ChrNumber)'_'(Position) for the window
        - 7: End (ChrNumber)'_'(Position) for the window
        - 8: identification of window: (ChrNumber)'_'(startPositionMBP)
        - 9: variance explained by n adjacents SNP



# Setup

```{r, warning = FALSE, message=FALSE}
snp_sol <-
  read_table2(
    here::here("data/derived_data/gwas_hair/iter4/chrsnp_pval"),
    col_names = c("trait", "effect", "pval", "snp", "chr", "pos")
  )  %>%
  # SNP effects & weights
  left_join(read_table2(here::here(
    "data/derived_data/gwas_hair/iter4/snp_sol"
  ))) %>%
  rename(neglog10p = pval) %>%
  select(-trait) %>%
  # Change -log(10) values back
  mutate(pval = 10 ^ (-(neglog10p))) %>%
  # Keep only autosomes
  filter(29 >= chr) %>%
  # Remove SNPs with no pvalue
  filter(!is.nan(pval)) %>%
  # allele frequency data
  left_join(read_table2(
    here::here(
      "data/derived_data/gwas_hair/iter4/freqdata.count.after.clean"
    ),
    col_names = c("snp", "af", "snp_comment")
  )) %>%
  mutate_at(vars("snp_effect", "weight"), ~ as.numeric(.))
  
```

```{r}
snp_sol %>% 
  filter(converted_effect > 100) %>% 
  select(chr, pos, snp_effect, variance_explained, var_a_hat, neglog10p, af)
```


# SNP effects

```{r, fig.width = 10, fig.height = 5.6}
snp_sol %>%
  filter(snp_effect < 700) %>% 
  hair_manhattan(
    y_var = abs(snp_effect),
    y_lab = "Absolute value of the estimated SNP effect",
    plot_title = NULL
  )
```


# P-values

```{r, fig.width = 10, fig.height = 5.6}
snp_sol %>%
  hair_manhattan(
    y_var = neglog10p,
    y_lab = "-log10(p-value)",
    plot_title = NULL,
    sigline = TRUE
  )
```

## qq-plot

```{r}
ggqq <- function(pvector){
  pvector = pvector[!is.na(pvector) & !is.nan(pvector) & !is.null(pvector) & is.finite(pvector) & pvector<1 & pvector>0]
  pdf = data.frame(observed = -log10(sort(pvector, decreasing = FALSE)), expected = -log10(ppoints(length(pvector))))
  qqplotted = ggplot(pdf, aes(expected, observed)) +
    geom_point() +
    geom_abline(intercept = 0,
                slope = 1,
                colour = "red")+
    labs(x = expression(paste("Expected ", -log10, "(", italic('p'), ")")),
         y = expression(paste("Expected ", -log10, "(", italic('p'), ")")))
  return(qqplotted)
}
```


```{r}
ggqq(pvector = snp_sol$pval)
```


# Commentary