---
title: "Weaning weight versus hair shedding covariances & genetic correlations"
author: "Harly Durbin"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cols.print = 6)
library(readr)
library(dplyr)
library(tidyr)
library(forcats)
library(stringr)
library(tibble)
library(glue)
library(rlang)
library(magrittr)
library(purrr)
library(tidylog)

source(here::here("source_functions/melt_aireml.R"))
source(here::here("source_functions/calculate_heritability.R"))
source(here::here("source_functions/lrt_calc.R"))
source(here::here("source_functions/parse_loglik.R"))

options(scipen=999)
```

# Notes & questions

In old results based on pedigree:

* Model 1 = `190916_ww`
* Model 2 = `190926_ww/all_effect`
    + Results not reported in the manuscript
* Model 3 = `190926_ww/no_direct`
    + Model 3 here is what I refer to as model 2 in the manuscript (since model 2 isn't reported)

---

# Setup 

```{r}
prettygen <- function(model, dataset = "all") {
  # This is lazy and I don't care
  path <- if (dataset == "all") {
    
    # This is lazy but I didn't anticipate running so many versions
    glue::glue(
      "data/derived_data/ww_genetic_corr/single_step/model{model}/airemlf90.model{model}.log"
    )
  } else if (dataset == "fescue_yes") {
    glue::glue(
      "data/derived_data/ww_fescue/model{model}_yes/airemlf90.model{model}_yes.log"
    )
  } else if (dataset == "fescue_no") {
    glue::glue(
      "data/derived_data/ww_fescue/model{model}_no/airemlf90.model{model}_no.log"
    )
  } else if (dataset == "ww_no") {
    glue::glue(
      "data/derived_data/ww_fescue/model{model}_no_ww/airemlf90.model{model}_no_ww.log"
    )
  } else if (dataset == "ww_yes") {
    glue::glue(
      "data/derived_data/ww_fescue/model{model}_yes_ww/airemlf90.model{model}_yes_ww.log"
    )
  } 
  
  read_table2(
    here::here(path),
    skip = 9,
    n_max = 4,
    col_names = c(
      glue("Hair shedding (direct)"),
      glue("Weaing weight (direct)"),
      glue("Hair shedding (maternal)"),
      glue("Weaning weight (milk)")
    )
  ) %>%
    mutate(rowname = colnames(.)) %>%
    column_to_rownames(var = "rowname") %>%
    DT::datatable(
      filter = 'none',
      extensions = list('FixedColumns', 'RowReorder'),
      options = list(
        fixedColumns = list(leftColumns = 1),
        rowReorder = FALSE
      )
    )
  
}
```

```{r}
prettyvar <- function(varcovdf) {
  varcovdf %>% 
    filter(val1 == val2) %>%
    separate(val1, into = c("Trait", "effect"), sep = "_") %>%
    select(Trait, effect, var_cov) %>%
    mutate(effect = str_to_upper(effect)) %>% 
    tidyr::pivot_wider(names_from = effect, values_from = var_cov) %>%
    rename(
      Direct = DIR,
      Maternal = MAT,
      Residual = RES
    ) %>%
    mutate(Trait = if_else(Trait == "hair", "Hair shedding", "Weaning weight"))
}
```

# Genetic correlations

## Model 1 (no maternal effect of hair shedding, weaning weight as trait of the calf) 

> Sample of data: relationship is between animal's own weaning weight and animal's own hair score

```{r}
read_table2(here::here("data/derived_data/ww_genetic_corr/single_step/model1/data.txt"), 
            col_names = c("reg_number", "hair_cg", "ww_cg", "hair_score", "ww")) %>% 
  sample_n(10)
```


```{r, cols.print = 6}
prettygen(model = 1)
```

## Model 2 (all effects, weaning weight as trait of the dam) 

```{r, cols.print = 6}
prettygen(model = 2)
```


## Model 3 ($HS_{calf}$ fixed to zero, weaning weight as trait of the dam) 

> Sample of data: relationship is between calf's weaning weight and its dam's hair score

```{r}
read_table2(here::here("data/derived_data/ww_genetic_corr/single_step/model3/data.txt"), 
            col_names = c("calf_reg", "hair_cg", "ww_cg", "dam_hair_score", "calf_ww")) %>% 
  sample_n(10)
```

### All possible data

```{r}
prettygen(model = 3)
```

> Residual correlations

```{r}

read_table2(
  here::here(
    "data/derived_data/ww_genetic_corr/single_step/model3/airemlf90.model3.log"
  ),
  skip = 35,
  n_max = 2,
  col_names = c(glue("Hair shedding"),
                glue("Weaing weight"))
) %>%
  mutate(rowname = colnames(.)) %>%
  column_to_rownames(var = "rowname") %>%
  DT::datatable(
    filter = 'none',
    extensions = list('FixedColumns', 'RowReorder'),
    options = list(
      fixedColumns = list(leftColumns = 1),
      rowReorder = FALSE
    )
  )

```


### Toxic fescue = NO

```{r}
prettygen(model = 3, dataset = "fescue_no")
```

#### For comparison, toxic fescue = NO weaning weight only

```{r}
read_table2(
  here::here("data/derived_data/ww_fescue/model3_no_ww/airemlf90.model3_no_ww.log"),
  skip = 7,
  n_max = 2,
  col_names = c(
    glue("Weaing weight (direct)"),
    glue("Weaning weight (milk)")
  )
) %>%
  mutate(rowname = colnames(.)) %>%
  column_to_rownames(var = "rowname") %>%
  DT::datatable(
    filter = 'none',
    extensions = list('FixedColumns', 'RowReorder'),
    options = list(
      fixedColumns = list(leftColumns = 1),
      rowReorder = FALSE
    )
  )

```


### Toxic fescue = YES

```{r}
prettygen(model = 3, dataset = "fescue_yes")
```

#### For comparison, toxic fescue = YES weaning weight only

```{r}
read_table2(
  here::here("data/derived_data/ww_fescue/model3_yes_ww/airemlf90.model3_yes_ww.log"),
  skip = 7,
  n_max = 2,
  col_names = c(
    glue("Weaing weight (direct)"),
    glue("Weaning weight (milk)")
  )
) %>%
  mutate(rowname = colnames(.)) %>%
  column_to_rownames(var = "rowname") %>%
  DT::datatable(
    filter = 'none',
    extensions = list('FixedColumns', 'RowReorder'),
    options = list(
      fixedColumns = list(leftColumns = 1),
      rowReorder = FALSE
    )
  )

```

# Variance components, heritability, and repeatability

## Model 1 (no maternal effect of hair shedding, weaning weight as trait of the calf) 

```{r}
varcov1 <-
  melt_aireml(
  path = here::here("data/derived_data/ww_genetic_corr/single_step/model1/airemlf90.model1.log"),
  effect2 = c("hair_dir", "ww_dir", "hair_mat", "ww_mat"),
  effect4 = c("hair_mpe", "ww_mpe"),
  effect5 = c("hair_pe", "ww_pe"),
  resids = c("hair_res", "ww_res")
)
```

```{r, cols.print = 6}
prettyvar(varcovdf = varcov1)
```

```{r}
herit1 <-
  varcov1 %>% 
  biv_heritability(
    abbrvs = c("hair", "ww"),
    descs = c("Hair shedding", "Weaning weight"),
    mat = TRUE,
    mpe = TRUE,
    pe = TRUE
  )
```

```{r}
herit1 %>% 
  rename(Trait = key) 
```

## Model 2 (all effects, weaning weight as trait of the dam) 

```{r}

varcov2 <-
melt_aireml(
  path = here::here("data/derived_data/ww_genetic_corr/single_step/model2/airemlf90.model2.log"),
  effect2 = c("hair_dir", "ww_dir", "hair_mat", "ww_mat"),
  effect4 = c("hair_mpe", "ww_mpe"),
  resids = c("hair_res", "ww_res")
) 
```



```{r, cols.print = 6}
prettyvar(varcovdf = varcov2)
```

```{r}
herit2 <-
  varcov2 %>% 
  biv_heritability(
    abbrvs = c("hair", "ww"),
    descs = c("Hair shedding", "Weaning weight"),
    mat = TRUE,
    mpe = TRUE
  )
```

```{r}
herit2 %>% 
  rename(Trait = key)
```


## Model 3 ($HS_{calf}$ fixed to zero, weaning weight as trait of the dam) 

### All possible data

```{r}
varcov3 <-
melt_aireml(
  path = here::here("data/derived_data/ww_genetic_corr/single_step/model3/airemlf90.model3.log"),
  effect2 = c("hair_dir", "ww_dir", "hair_mat", "ww_mat"),
  effect4 = c("hair_mpe", "ww_mpe"),
  resids = c("hair_res", "ww_res")
)
```


```{r, cols.print = 6}
prettyvar(varcovdf = varcov3)
```

```{r}
herit3 <-
  varcov3 %>% 
  biv_heritability(
    abbrvs = c("hair", "ww"),
    descs = c("Hair shedding", "Weaning weight"),
    mat = TRUE,
    mpe = TRUE
  )
```

```{r}
herit3 %>% 
  rename(Trait = key) 
```

### Toxic fescue = NO

```{r}
varcov3_no <-
melt_aireml(
  path = here::here("data/derived_data/ww_fescue/model3_no/airemlf90.model3_no.log"),
  effect2 = c("hair_dir", "ww_dir", "hair_mat", "ww_mat"),
  effect4 = c("hair_mpe", "ww_mpe"),
  resids = c("hair_res", "ww_res")
)
```


```{r, cols.print = 6}
prettyvar(varcovdf = varcov3_no)
```

```{r}
herit3_no <-
  varcov3_no %>% 
  biv_heritability(
    abbrvs = c("hair", "ww"),
    descs = c("Hair shedding", "Weaning weight"),
    mat = TRUE,
    mpe = TRUE
  )
```

```{r}
herit3_no %>% 
  rename(Trait = key) 
```

#### For comparison, toxic fescue = NO weaning weight only

```{r}
varcov3_no_ww <-
melt_aireml(
  path = here::here("data/derived_data/ww_fescue/model3_no_ww/airemlf90.model3_no_ww.log"),
  effect2 = c("ww_dir", "ww_mat"),
  effect4 = c("ww_mpe"),
  resids = c("ww_res")
)
```

```{r, cols.print = 6}
prettyvar(varcovdf = varcov3_no_ww)
```

```{r}
herit3_no_ww <-
  varcov3_no_ww %>% 
  univ_heritability(
    abbrv = c("ww"),
    desc = c("Weaning weight"),
    mat = TRUE,
    mpe = TRUE
  )
```

```{r}
herit3_no_ww %>% 
  rename(Trait = key) 
```

### Toxic fescue = YES

```{r}
varcov3_yes <-
melt_aireml(
  path = here::here("data/derived_data/ww_fescue/model3_yes/airemlf90.model3_yes.log"),
  effect2 = c("hair_dir", "ww_dir", "hair_mat", "ww_mat"),
  effect4 = c("hair_mpe", "ww_mpe"),
  resids = c("hair_res", "ww_res")
)
```

```{r, cols.print = 6}
prettyvar(varcovdf = varcov3_yes)
```

```{r}
herit3_yes <-
  varcov3_yes %>% 
  biv_heritability(
    abbrvs = c("hair", "ww"),
    descs = c("Hair shedding", "Weaning weight"),
    mat = TRUE,
    mpe = TRUE
  )
```

```{r}
herit3_yes %>% 
  rename(Trait = key) 
```

#### For comparison, toxic fescue = YES weaning weight only

```{r}
varcov3_yes_ww <-
melt_aireml(
  path = here::here("data/derived_data/ww_fescue/model3_yes_ww/airemlf90.model3_yes_ww.log"),
  effect2 = c("ww_dir", "ww_mat"),
  effect4 = c("ww_mpe"),
  resids = c("ww_res")
)
```

```{r, cols.print = 6}
prettyvar(varcovdf = varcov3_yes_ww)
```

```{r}
herit3_yes_ww <-
  varcov3_yes_ww %>% 
  univ_heritability(
    abbrv = c("ww"),
    desc = c("Weaning weight"),
    mat = TRUE,
    mpe = TRUE
  )
```

```{r}
herit3_yes_ww %>% 
  rename(Trait = key) 
```
# Likelihood ratio test

* Test for model fit between including vs. excluding calf effect on hair shedding

```{r}
no_direct <- parse_loglik(path = here::here("data/derived_data/ww_genetic_corr/pedigree/model3/airemlf90.model3.log"), option = "loglik")

all_effects <- parse_loglik(path = here::here("data/derived_data/ww_genetic_corr/pedigree/model2/airemlf90.model2.log"), option = "loglik")

lrt_calc(ll_null = no_direct, ll_test = all_effects)
```

# Commentary

```{r}
wean_dat <- read_table2(here::here("data/derived_data/ww_genetic_corr/single_step/model3/data.txt"), col_names = c("full_reg", "hair_cg", "ww_cg", "hair_score", "ww"))
```


Phenotypes and contemporary group designations came from from weekly growth run of the American Angus Association National Cattle Evaluation (NCE). Prior to entering the NCE, phenotypes were adjusted for age of dam and to 205 days of age. Data were aggregated for cows with at least one hair shedding score recorded, all of their recorded calves, their weaning weight contemporary group peers, and all of their recorded calves' weaning weight contemporary group peers. Phenotypes from calves born via embryo transfer and contemporary groups with fewer than five animals or no variation were excluded, resulting in `r wean_dat %>% filter(ww != 0) %>% select(full_reg) %>% n_distinct(.) %>% scales::comma()` total weaning weight and `r wean_dat %>% filter(hair_score != 0) %>% select(full_reg) %>% n_distinct(.) %>% scales::comma()` total hair shedding records. Of animals retained for analysis, `r wean_dat %>% filter(hair_score != 0) %>% filter(ww != 0) %>% select(full_reg) %>% n_distinct(.) %>% scales::comma()` had both a recorded weaning weight and at least one hair shedding score.

In model 1, hair shedding was fit as a trait of the dam and weaning weight as a trait of the calf. The effects fit in model 1 were $HS_{dam}$, $WW_{calf}$, $WW_{maternal}$, $WW_{MPE}$, $HS_{PE}$ and their covariances.  

$$y_t = Xb + Z_1u + Z_2pe + Z_3m + Z_4mpe + e$$

Where:

* $t$ the trait (hair shedding score or weaning weight)
* $y$ is phenotype
* $b$ is contemporary group
* $u$ is additive genetic effect
* $p$ is permanent environment effect
* $m$ is maternal genetic effect
* $mpe$ is maternal permanent environment
* $e$ is random residual

Conceivably, environmental factors affecting a dam's hair shedding performance could also affect the direct weaning weight of her calf, creating a residual covariance between the two traits. In order to model this covariance, two subsequent models were fit in which both hair shedding and the weaning weight of the calf she raised during the scoring year were treated as traits of the dam.    

For cows with hair shedding scores but no calf weaning weight reported during the scoring year, a "dummy calf" with a weaning weight set to missing was created.  

In model 3, weaning weight was also treated as a trait of the dam, but the effect of the calf on the dam's hair shedding score was fixed to zero. The effects fit in model 3 were: 

* $WW_{calf}$
* $WW_{maternal}$
* $HS_{maternal}$
* $WW_{MPE}$
* $HS_{PE}$

$$y_t = Xb + Z_1u + Z_2pe + Z_3m + Z_4mpe + e$$

Where:

* $t$ the trait (hair shedding score or weaning weight)
* $y$ is phenotype
* $b$ is contemporary group
* $u$ is additive genetic effect (fit only for weaning weight)
* $p$ is permanent environment effect (fit only for hair shedding)
* $m$ is maternal genetic effect 
* $mpe$ is maternal permanent environment (fit only for weaning weight)
* $e$ is random residual

---

Hair shedding score heritability estimates were similar for all models, ranging from `r min(as.numeric(herit1[1,3]), as.numeric(herit2[1,4]), as.numeric(herit3[1,4]))` to `r max(as.numeric(herit1[1,3]), as.numeric(herit2[1,4]), as.numeric(herit3[1,4]))`.  


Models 2 and 3 created a direct tie between a dam's hair shedding score and the calf she weaned that year.  

---

The effects fit in model 2 were:  

* $HS_{calf}$
* $WW_{calf}$
* $WW_{maternal}$
* $HS_{maternal}$
* $WW_{MPE}$
* $HS_{PE}$

and their covariances. A residual covariance was fit between hair shedding and weaning weight.

